BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-02-28 paper/paper.qmd
First pass of constraints and start on projection vs rescaling section

@@

### Sub-block Descent
- Use smaller blocks, but descent in parallel (sub-blocks don't wait for other sub-blocks)
- Can perform this efficiently with a "matrix step-size"

@@

@@

### Momentum
- This one is standard
- Use something similar to [@xu_BlockCoordinateDescent_2013]
- This is compatible with sub-block descent with appropriately defined matrix operations

@@

@@

### Constraints
- one type of update (other than the typical GD update)
- can combine them with composition
  - which is different than projecting onto their intersection!
- Constraint updates combine the constraint with how they are enforced
  - need to go together since there are multiple ways to enforce them e.g. simplex (see next section)

# Partial Projection and Rescaling {#sec-ppr}
- for bounded linear constraints
  - first project
  - then rescale to enforce linear constraints
- faster to execute then a projection
- often does not loose progress because of the rescaling (decomposition dependent)

>>>>

For a vector, matrix, or tensor filled with ones, we use $\mathbb{1}\in\mathbb{R}^{I_1 \times \dots \times I_N}$. This can be constructed in Julia with `ones(I₁, ..., Iₙ)`.

@@

### Sub-block Descent {#sec-sub-block-descent}
- Use smaller blocks, but descent in parallel (sub-blocks don't wait for other sub-blocks)
- Can perform this efficiently with a "matrix step-size"

@@

TODO Compare to preconditioned decent. See [@kunstner_searching_2023;@gao_gradient_2024;@gao_scalable_2024;@qu_optimal_2024]

@@

### Momentum {#sec-momentum}
- This one is standard
- Use something similar to [@xu_BlockCoordinateDescent_2013]
- This is compatible with sub-block descent with appropriately defined matrix operations

@@

### Constraints {#sec-constraints}
- one type of update (other than the typical GD update)
- can combine them with composition
  - which is different than projecting onto their intersection!
- Constraint updates combine the constraint with how they are enforced
  - need to go together since there are multiple ways to enforce them e.g. simplex (see next section)

One of the main motivations for developing BlockTensorDecomposition.jl is to solve constrained tensor problems. Other code did not have the expressivity to handle constraints beyond the most common constraints on the factors: nonnegativity and Euclidean normalized columns. To enable flexibility, BlockTensorDecomposition.jl defines

```julia
abstract type AbstractConstraint <: Function end
```

which has two interfaces. The first treats the constraint like a function

```julia
(C::AbstractConstraint)(A::AbstractArray)
```

and applies the constraint to an abstract array, and the other interface

```julia
check(C::AbstractConstraint, A::AbstractArray)
```

checks if the array `A` satisfies the constraint `C`. A generic constrain only needs to define these two functions.

```julia
struct GenericConstraint <: AbstractConstraint
    apply::Function # input a AbstractArray -> mutate it so that `check` would return true
    check::Function
end

function (C::GenericConstraint)(D::AbstractArray)
    (C.apply)(D)
end

check(C::GenericConstraint, A::AbstractArray) = (C.check)(A)
```

In general, the function `check` could be defined as the following,

```julia
function check(C, A)
    A_copy = copy(A)
    C(A)
    return A ≈ A_copy
end
```

but there is often an easier way to check if a tensor is constrained than to apply the constraint.

Although our basic block projected gradient descent algorithm @eq-proximal-explicit relies on Euclidean projections to the relevant constraint set, we want to remain flexible and allow for other maps that move an iterate to the constraint set. So each constraint needs to store more than just the constraint itself (the *what*), but also the map from an iterate to the constraint set (the *how*). When prototyping algorithms, it is worth comparing alternate approaches to see if there is a more efficient method to enforce a given constraint (See @sec-ppr for an example of constraining to a set without Euclidean projections).

The first class of constraints are entry-wise constraints. This is convenient for defining a scalar function that gets applied to all entries of a tensor.

```julia
struct Entrywise <: AbstractConstraint
    apply::Function
    check::Function
end

function (C::Entrywise)(A::AbstractArray)
    A .= (C.apply).(A)
end

check(C::Entrywise, A::AbstractArray) = all((C.check).(A))
```

Some common examples would be a nonnegative constraint, constraining entries to an interval, or constraining entries to be one or zero.

```julia
nonnegative! = Entrywise(x -> max(0, x), x ≥ 0)

IntervalConstraint(a, b) = Entrywise(x -> clamp(x, a, b), x -> a ≤ x ≤ b)

binary! = Entrywise(x > 0.5 ? one(x) : zero(x), x -> x in (0, 1))
```

Another class of constraints are normalizations $\mathcal{C}_{\lVert \cdot \rVert_{a}}=\left\{v \middle| \lVert v \rVert_{a} = 1 \right\}$ for some norm $\lVert \cdot \rVert_{a}$. These can be enforced by a (Euclidean) projection

$$
\hat{v} \in \arg\,\min_{u\in\mathcal{C}_{\lVert \cdot \rVert_a}} \lVert u - v \rVert _2^2,
$$

or a scaling (assuming $v\neq 0$)

$$
\hat{v} \leftarrow \frac{v}{\lVert v \rVert_a}.
$$

These operations agree for the Frobenius norm (entry-wise $2$-norm), but are different operations in general. In BlockTensorDecomposition.jl, we define these classes as the following.

TODO simplify the following code?

```julia
abstract type AbstractNormalization <: AbstractConstraint end
```

```julia
struct ProjectedNormalization <: AbstractNormalization
    norm::Function # calculate the norm of an AbstractArray
    projection::Function # mutate an AbstractArray so it has norm == 1
    whats_normalized::Function # what part of the array is normalized
                               # e.g. eachrow, or omit for the entire array
end

ProjectedNormalization(norm, projection; whats_normalized=identityslice) =
    ProjectedNormalization(norm, projection, whats_normalized)

function (P::ProjectedNormalization)(A::AbstractArray)
    whats_normalized_A = P.whats_normalized(A)
    (P.projection).(whats_normalized_A)
end

check(P::ProjectedNormalization, A::AbstractArray) =
    all((P.norm).(P.whats_normalized(A)) .≈ 1)
```

```julia
struct ScaledNormalization{T<:Union{Real,AbstractArray{<:Real},Function}} <: AbstractNormalization
    norm::Function # calculate the norm of an AbstractArray
    whats_normalized::Function # what part of the array is normalized
    scale::T # the scale that `whats_normalized` should be normalized to
             # Can be a real, array of reals, or function of the input that
             # that returns a real or array of reals
end

ScaledNormalization(norm;whats_normalized=identityslice,scale=1) =
    ScaledNormalization{typeof(scale)}(norm, whats_normalized, scale)

function (S::ScaledNormalization{T})(A::AbstractArray) where {T<:Union{Real,AbstractArray{<:Real}}}
    whats_normalized_A = S.whats_normalized(A)
    A_norm = (S.norm).(whats_normalized_A) ./ S.scale
    whats_normalized_A ./= A_norm
    return A_norm
end

function (S::ScaledNormalization{T})(A::AbstractArray) where {T<:Function}
    whats_normalized_A = S.whats_normalized(A)
    A_norm = (S.norm).(whats_normalized_A) ./ S.scale(A)
    whats_normalized_A ./= A_norm
    return A_norm
end

check(S::ScaledNormalization{<:Union{Real,AbstractArray{<:Real}}}, A::AbstractArray) =
    all((S.norm).(S.whats_normalized(A)) .≈ S.scale)

check(S::ScaledNormalization{<:Function}, A::AbstractArray) =
    all((S.norm).(S.whats_normalized(A)) .≈ S.scale(A))
```

We can define some common constraints like $L_1$ normalized through a projection

```julia
l1normalize! =
    ProjectedNormalization(l1norm, l1project!)
l1normalize_rows! =
    ProjectedNormalization(l1norm, l1project!; whats_normalized=eachrow)
l1normalize_1slices! =
    ProjectedNormalization(l1norm, l1project!;
        whats_normalized=(x -> eachslice(x; dims=1)))
```

or $L_\infty$ normalized by scaling

```julia
linftyscale_cols! = ScaledNormalization(linftynorm; whats_normalized=eachcol)
```

or even normalize the 3-fibres of a third order tensor on average!

```julia
l2scale_average12slices! = ScaledNormalization(l2norm;
    whats_normalized=(x -> eachslice(x; dims=1)),
    scale=(A -> size(A, 2)))
```

Constraints can also be composed. This is *not* the same as the intersection of constraints. This is only a handy way to apply multiple constraints in series, but there is no clever logic that interprets the constraints and combines them into a single constraint.

```julia
struct ComposedConstraint{T<:AbstractConstraint, U<:AbstractConstraint} <: AbstractConstraint
    outer::T
    inner::U
end

function (C::ComposedConstraint)(A::AbstractArray)
    C.inner(A)
    C.outer(A)
end

check(C::ComposedConstraint, A::AbstractArray) =
    check(C.outer, A) & check(C.inner, A)

Base.:∘(f::AbstractConstraint, g::AbstractConstraint) =
    ComposedConstraint(f, g)
```

This means the following three constraints are all different!

```julia
l1normalize! ∘ nonnegative!
nonnegative! ∘ l1normalize!
simplex!
```

See @sec-ppr for a discussion on why it may be advantages to use one of these constraints over the other.

@@

To put all these pieces together, we define an abstract type that can be subtyped for the various types of update. This will include gradient descent steps, momentum updates, and constraint enforcing updates.

```julia
abstract type AbstractUpdate <: Function end
```

A generic update is a function that can take some keyword arguments and mutate an iterate.

```julia
struct GenericUpdate <: AbstractUpdate
    f::Function
end

(U::GenericUpdate)(x; kwargs...) = U.f(x; kwargs...)
```

The first types of updates are gradient descent updates.

```julia
abstract type AbstractGradientDescent <: AbstractUpdate end
```

This includes the regular gradient descent update, and also the sub-block descent updates.

```julia
struct GradientDescent <: AbstractGradientDescent
    n::Integer
    gradient::Function
    step::AbstractStep
end

function (U::GradientDescent)(x; x_last, kwargs...)
    n = U.n
    if checkfrozen(x, n)
        return x
    end
    grad = U.gradient(x; kwargs...)
    # Note we pass a function for grad_last (lazy) so that we only compute it if needed for the step
    s = U.step(x; n, x_last, grad, grad_last=(x -> U.gradient(x; kwargs...)), kwargs...)
    a = factor(x, n)
    @. a -= s*grad
end
```

The only addition to `BlockGradientDescent` is a function that combines the step with the gradient.

```julia
struct BlockGradientDescent <: AbstractGradientDescent
    n::Integer
    gradient::Function
    step::AbstractStep
    combine::Function # takes a step (number, matrix, or tensor) and combines it with a gradient
end

function (U::BlockGradientDescent)(x; x_last, kwargs...)
    n = U.n
    if checkfrozen(x, n)
        return x
    end
    grad = U.gradient(x; kwargs...)
    # Note we pass a function for grad_last (lazy) so that we only compute it if needed for the step
    s = U.step(x; n, x_last, grad, grad_last=(x -> U.gradient(x; kwargs...)), kwargs...)
    a = factor(x, n)
    a .-= U.combine(grad, s)
end
```

A separate step type is defined to allow for different types of steps like constant step, Lipschitz, and spectral projected gradient (SPG).

```julia
abstract type AbstractStep <: Function end
```

```julia
struct LipschitzStep <: AbstractStep
    lipschitz::Function
end

function (step::LipschitzStep)(x; kwargs...)
    L = step.lipschitz(x)
    return L^(-1) # allow for Lipschitz to be a diagonal matrix
end

function (step::LipschitzStep)(x::Tucker; kwargs...)
    L = step.lipschitz(x)
    if typeof(L) <: Tuple # Currently the only case is when we are updating the core of a Tucker factorization
                          # Using this condition as a way to tell if it is the core we are calculating the constant for
        return map(X -> X^(-1), L)
    else
        return L^(-1) # allow for Lipschitz to be a diagonal matrix
    end
end
```

```julia
struct ConstantStep <: AbstractStep
    stepsize::Real
end

(step::ConstantStep)(x; kwargs...) = step.stepsize
```

```julia
struct SPGStep <: AbstractStep
    min::Real
    max::Real
end

SPGStep(;min=1e-10, max=1e10) = SPGStep(min, max)

# Convert an input of the full decomposition, to a calculation on the nth factor
(step::SPGStep)(x::T; n, x_last::T, grad_last::Function, kwargs...) where {T <: AbstractDecomposition} =
    step(factor(x,n); x_last=factor(x_last,n), grad_last=grad_last(x_last), kwargs...)

function (step::SPGStep)(x; grad, x_last, grad_last, stepmin=step.min, stepmax=step.max, kwargs...)
    s = x - x_last
    y = grad - grad_last
    sy = (s ⋅ y)
    if sy <=0
        return stepmax
    else
        suggested_step = (s ⋅ s) / sy
        return clamp(suggested_step, stepmin, stepmax)
    end
end
```

The gradient and Lipschitz stepsize is calculated by a separate function that gets make on initialization of the `factorization` algorithm (See @sec-gradient-computation and @sec-lipschitz-computation). This de-couples applying the gradient descent (the *what*), and the computation of the gradient (the *how*) so that the gradient can be calculated manually (if an efficient method is known), or with automatic differentiation. This also allows other updates to use the same computation code. For example, momentum updates also use the same Lipschitz calculation function. Momentum updates also use the same `combine` function as the sub-block gradient descent updates.

```julia
struct MomentumUpdate <: AbstractUpdate
    n::Integer
    lipschitz::Function
    combine::Function # How to combine the momentum variable `ω` with a factor `a`
end

MomentumUpdate(n, lipschitz) = MomentumUpdate(n, lipschitz, (ω, a) -> ω * a)

function MomentumUpdate(GD::AbstractGradientDescent)
    n, step = GD.n, GD.step
    @assert typeof(step) <: LipschitzStep

    return MomentumUpdate(n, step.lipschitz)
end

function MomentumUpdate(GD::BlockGradientDescent)
    n, step, combine = GD.n, GD.step, GD.combine
    @assert typeof(step) <: LipschitzStep

    return MomentumUpdate(n, step.lipschitz, combine)
end
```

Constraints also get their own abstract subtype of updates.

```julia
abstract type ConstraintUpdate <: AbstractUpdate end
```

We define a constructor for applying a constraint to a particular factor $n$. This turns an `AbstractConstraint` into a `ConstraintUpdate`.

```julia
ConstraintUpdate(n, constraint::GenericConstraint; kwargs...) =
    GenericConstraintUpdate(n, constraint)
ConstraintUpdate(n, constraint::ProjectedNormalization; kwargs...) =
    Projection(n, constraint)
ConstraintUpdate(n, constraint::Entrywise; kwargs...) =
    Projection(n, constraint)

function ConstraintUpdate(n, constraint::ScaledNormalization; skip_rescale=false, whats_rescaled=missing, kwargs...)
    if skip_rescale
        ismissing(whats_rescaled) ||
        isnothing(whats_rescaled) ||
        @warn "skip_rescale=true but whats_rescaled=$whats_rescaled was given. Overriding to whats_rescaled=nothing"
        return Rescale(n, constraint, nothing)
    else
        return Rescale(n, constraint, whats_rescaled)
    end
end
```

A generic constraint update extracts the factor it needs to constrain, and applies the constraint to that factor.

```julia
struct GenericConstraintUpdate <: ConstraintUpdate
    n::Integer
    constraint::GenericConstraint
end

check(U::GenericConstraintUpdate, D::AbstractDecomposition) = check(U.constraint, factor(D, U.n))

function (U::GenericConstraintUpdate)(x::T; kwargs...) where T
    n = U.n
    A = factor(x, n)
    U.constraint(A)
    check(U, A) ||
        error("Something went wrong with GenericConstraintUpdate: $GenericConstraintUpdate")
end
```

Projections follow this pattern closely.

```julia
struct Projection <: ConstraintUpdate
    n::Integer
    proj::Union{ProjectedNormalization, Entrywise}
end

check(P::Projection, D::AbstractDecomposition) = check(P.proj, factor(D, P.n))

function (U::Projection)(x::T; kwargs...) where T
    n = U.n
    U.proj(factor(x, n))
end
```

A more involved type of constraint update is a rescaled update. This uses a scaled normalization, but moves the weight of the factor to other factors.

For example, if we have three matrix factors $A, B, C$ in a CP decomposition where we want the sum of the entries in $A$ to sum to $1$, we can divide $A$ by its sum, and multiple factor $B$ by this amount. That way the recombined tensor $\left\llbracket A, B, C \right\rrbracket$ remains unchanged, but now $A$ satisfies the desired constraint. We could instead multiply both $B$ and $C$ by the square root of the sum to achieve a similar outcome. When it is not specified what is rescaled (`whats_rescaled = missing`), we assume we should multiply every other factor by the geometric mean of the scaling. If we just want to scale the factor but skip any rescaling, we can use `whats_rescaled = nothing`.

```julia
struct Rescale{T<:Union{Nothing,Missing,Function}} <: ConstraintUpdate
    n::Integer
    scale::ScaledNormalization
    whats_rescaled::T
end

check(S::Rescale, D::AbstractDecomposition) = check(S.scale, factor(D, S.n))

function (U::Rescale{<:Function})(x; kwargs...)
    Fn_scale = U.scale(factor(x, U.n))
    to_scale = U.whats_rescaled(x)
    to_scale .*= Fn_scale
end

(U::Rescale{Nothing})(x; kwargs...) =
    U.scale(factor(x, U.n))

function (U::Rescale{Missing})(x; skip_rescale=false, kwargs...)
    Fn_scale = U.scale(factor(x, U.n))
    x_factors = factors(x)
    N = length(x_factors) - 1

    # Nothing to rescale, so return here
    if N == 0 || skip_rescale
        return nothing
    end

    # Assume we want to evenly rescale all other factors by the Nth root of Fn_scale
    scale = geomean(Fn_scale)^(1/N)
    for (i, A) in zip(eachfactorindex(x), x_factors)
        # skip over the factor we just updated
        if i == U.n
            continue
        end
        A .*= scale
    end
end
```

See @sec-ppr for a more in-depth discussion on when it may be beneficial to use this type of constraint update over a simple projection.

We would like to combine all these updates to execute them in serial. We use a `BlockedUpdate` type to do this. This should be thought of as a list of `AbstractUpdates` that get applied one after another.

```julia
struct BlockedUpdate <: AbstractUpdate
    updates::Vector{AbstractUpdate}
end
```

::: {.callout-note}
# Technical Julia Note

We need the updates to be exactly something of the form `AbstractUpdate[]` since we want to push any type of `AbstractUpdate`s such as a `MomentumUpdate` or another `BlockedUpdate`, even if not already present. This means it cannot be `Vector{<:AbstractUpdate}` since a `BlockedUpdate` constructed with only `GradientDescent` would give a `GradientDescent[]` vector and we couldn't push a `MomentumUpdate`. And it cannot be `AbstractVector{AbstractUpdate}` since we may not be able to `insert!` or `push!` into other `AbstractVectors` like `Views`.
:::

We forward many standard methods so that `BlockedUpdate`s can behave like usual Julia vectors.

TODO should I actually show all these details?

```julia
Base.getindex(U::BlockedUpdate, i::Int) = getindex(updates(U), i)
Base.getindex(U::BlockedUpdate, I::Vararg{Int}) = getindex(updates(U), I...)
Base.getindex(U::BlockedUpdate, I) = getindex(updates(U), I) # catch all
Base.firstindex(U::BlockedUpdate) = firstindex(updates(U))
Base.lastindex(U::BlockedUpdate) = lastindex(updates(U))
Base.keys(U::BlockedUpdate) = keys(updates(U))
Base.length(U::BlockedUpdate) = length(updates(U))
Base.iterate(U::BlockedUpdate, state=1) = state > length(U) ? nothing : (U[state], state+1)
Base.filter(f, U::BlockedUpdate) = BlockedUpdate(filter(f, updates(U)))
```

We define how `BlockedUpdates` get applied to a tensor with the following function.

```julia
function (U::BlockedUpdate)(x::T; recursive_random_order::Bool=false, random_order::Bool=recursive_random_order, kwargs...) where T
    U_updates = updates(U)
    if random_order
        order = shuffle(eachindex(U_updates))
        U_updates = U_updates[order]
    end

    for update! in U_updates
        update!(x; recursive_random_order, kwargs...)
        # note random_order does not get passed down
    end
end
```

The default order the blocks are updated is cyclically through each factor of the decomposition `D::AbstractDecomposition`, in the order of `factors(D)`. For `AbstractTucker` decompositions like Tucker, Tucker-1, and CP, this means starting with the core, followed by the matrix factor for the first dimension, second dimension, and so on.

As an example, this would be the default order of updates for nonnegative CP decomposition on an order 3 tensor.

```julia
BlockedUpdate(
    MomentumUpdate(1, lipschitz)
    GradientStep(1, gradient, LipschitzStep)
    Projection(1, Entrywise(ReLU, isnonnegative))
    MomentumUpdate(2, lipschitz)
    GradientStep(2, gradient, LipschitzStep)
    Projection(2, Entrywise(ReLU, isnonnegative))
    MomentumUpdate(3, lipschitz)
    GradientStep(3, gradient, LipschitzStep)
    Projection(3, Entrywise(ReLU, isnonnegative))
)
```

The order of updates can be randomized with the `random_order` keyword.
```julia
X, stats, kwargs = factorize(Y; random_order=true)
```

By default, this will keep momentum steps, gradient steps, and constraint steps for each factor together as a block, in this order.

A possible order of updates could be the following. Note that the updates for each factor are grouped together, but each factor is updated in a random order.

```julia
BlockedUpdate(
    BlockedUpdate(
        MomentumUpdate(2, lipschitz)
        GradientStep(2, gradient, LipschitzStep)
        Projection(2, Entrywise(ReLU, isnonnegative))
    )
    BlockedUpdate(
        MomentumUpdate(1, lipschitz)
        GradientStep(1, gradient, LipschitzStep)
        Projection(1, Entrywise(ReLU, isnonnegative))
    )
    BlockedUpdate(
        MomentumUpdate(3, lipschitz)
        GradientStep(3, gradient, LipschitzStep)
        Projection(3, Entrywise(ReLU, isnonnegative))
    )
)
```

For more randomization, use the `recursive_random_order` keyword which will also randomize the order in which the momentum steps, gradient steps, and constraint steps are performed.

```julia
X, stats, kwargs = factorize(Y; recursive_random_order=true)
```

A possible order of updates could now be the following. The updates for each factor are still grouped together, but the updates within each block appear in a random order.

```julia
BlockedUpdate(
    BlockedUpdate(
        Projection(2, Entrywise(ReLU, isnonnegative))
        MomentumUpdate(2, lipschitz)
        GradientStep(2, gradient, LipschitzStep)
    )
    BlockedUpdate(
        MomentumUpdate(1, lipschitz)
        Projection(1, Entrywise(ReLU, isnonnegative))
        GradientStep(1, gradient, LipschitzStep)
    )
    BlockedUpdate(
        GradientStep(3, gradient, LipschitzStep)
        Projection(3, Entrywise(ReLU, isnonnegative))
        MomentumUpdate(3, lipschitz)
    )
)
```

The opposite of this would be to keep the outer order of blocks as given, but randomize the order which the updates for each factor gets applied, use the following code.

```julia
X, stats, kwargs = factorize(Y; recursive_random_order=true, random_order=false, group_by_factor=true)
```

A possible order of updates could now be the following. Note the order of factors is preserved (1, 2, 3) but the inner `BlockedUpdate`s have a random order.

```julia
BlockedUpdate(
    BlockedUpdate(
        Projection(1, Entrywise(ReLU, isnonnegative))
        MomentumUpdate(1, lipschitz)
        GradientStep(1, gradient, LipschitzStep)
    )
    BlockedUpdate(
        MomentumUpdate(2, lipschitz)
        Projection(2, Entrywise(ReLU, isnonnegative))
        GradientStep(2, gradient, LipschitzStep)
    )
    BlockedUpdate(
        GradientStep(3, gradient, LipschitzStep)
        MomentumUpdate(3, lipschitz)
        Projection(3, Entrywise(ReLU, isnonnegative))
    )
)
```

Note all the previously mentioned options still keeps the various updates for each factor together. For full randomization, use the following code.

```julia
X, stats, kwargs = factorize(Y; recursive_random_order=true, group_by_factor=false)
```

A possible order of updates could now be the following. Note that every update can appear anywhere in the order.

```julia
BlockedUpdate(
    Projection(3, Entrywise(ReLU, isnonnegative))
    MomentumUpdate(2, lipschitz)
    GradientStep(2, gradient, LipschitzStep)
    MomentumUpdate(1, lipschitz)
    GradientStep(1, gradient, LipschitzStep)
    Projection(2, Entrywise(ReLU, isnonnegative))
    MomentumUpdate(3, lipschitz)
    MomentumUpdate(2, lipschitz)
    Projection(1, Entrywise(ReLU, isnonnegative))
    GradientStep(3, gradient, LipschitzStep)
)
```

The complete behaviour is summarized in @tbl-blockupdate-randomization.

We also use `BlockedUpdate` to handle a composition of constraints.

<!-- ```julia
function ConstraintUpdate(n, constraint::ComposedConstraint; kwargs...)
    if constraint.inner == nonnegative! && typeof(constraint.outer) <: ScaledNormalization
        norm = constraint.outer.norm
        return BlockedUpdate(
            SafeNNProjection(n, ProjectedNormalization(norm, makeNNprojection(norm); whats_normalized=constraint.outer.whats_normalized)),
            ConstraintUpdate(n, constraint.outer; kwargs...))
    else
        return BlockedUpdate(ConstraintUpdate(n, constraint.inner; kwargs...), ConstraintUpdate(n, constraint.outer; kwargs...)) # note we apply inner constraint first
    end
end
``` -->

```julia
ConstraintUpdate(n, constraint::ComposedConstraint; kwargs...) =
    BlockedUpdate(ConstraintUpdate(n, constraint.inner; kwargs...),
                  ConstraintUpdate(n, constraint.outer; kwargs...))
end
```

The `BlockUpdate` language becomes especially helpful for automaically inserting additional updates as they are requested. For example, if we want to add momentum to a list of updates, we can call the following function.

```julia
function add_momentum!(U::BlockedUpdate)
    # Find all the GradientDescent updates
    U_updates = updates(U)
    indexes = findall(u -> typeof(u) <: AbstractGradientDescent, U_updates)

    # insert MomentumUpdates before each GradientDescent
    # do this in reverse order so "i" correctly indexes a GradientDescent
    # as we mutate updates
    for i in reverse(indexes)
        insert!(U_updates, i, MomentumUpdate(U_updates[i]))
    end
end
```

We can also interlace two lists of updates to put updates on the same factor next to each other, or group all updates by the factor they act on.

```julia
function smart_interlace!(U::BlockedUpdate, other_updates)
    for V in other_updates
        smart_insert!(U::BlockedUpdate, V::AbstractUpdate)
    end
end

smart_interlace!(U::BlockedUpdate, V::BlockedUpdate) = smart_interlace!(U::BlockedUpdate, updates(V))

function smart_insert!(U::BlockedUpdate, V::AbstractUpdate)
    U_updates = updates(U)
    i = findlast(u -> u.n == V.n, U_updates)

    # insert the other update immediately after
    # or if there is no update, push it to the end
    isnothing(i) ? push!(U_updates, V) : insert!(U_updates, i+1, V)
end
```

```julia
function group_by_factor(blockedupdate::BlockedUpdate)
    factor_labels = unique(getproperty(U, :n) for U in blockedupdate)
    updates_by_factor = [filter(U -> U.n == n, blockedupdate) for n in factor_labels]
    return BlockedUpdate(updates_by_factor)
end
```

# Rescaling to Constrain Tensor Factorization {#sec-ppr}
- for bounded linear constraints
  - first project
  - then rescale to enforce linear constraints
- faster to execute then a projection
- often does not loose progress because of the rescaling (decomposition dependent)

Constraints on factors in a tensor decomposition can arise naturally when modeling physical problems. A common class of constraints is normalizations which restrict a factor or slices of a factor to have unit norm. These are sometimes intersected with interval constraints such as requiring entries to be nonnegative.

For example, the demixing of $R$ probibility densities $b_1, \dots, b_R$ from $I$ mixtures $y_1,\dots,y_I$ for $I>R$ can be accomplished with a nonneagive matrix factorization (cite). We have the system of equations

$$
\begin{align}
{\ y_1} &= a_{11} {\ b_1} + a_{12} {\ b_2} + \dots + a_{1R} {\ b_R} \\
{\ y_2} &= a_{21} {\ b_1} + a_{22} {\ b_2} + \dots + a_{2R} {\ b_R} \\
            &\vdots \\
{\ y_I} &= a_{I1} {\ b_1} + a_{I2} {\ b_2} + \dots + a_{IR} {\ b_R} \\
\end{align}
$$

with unknown mixing coefficients $a_{i,r}$ and densities $b_r$. If we can discritized the mixtures $y_i$, we can rewrite this system as a rank $R$ factorization of the matrix $Y$ where $Y[i,:]=y_i$ and

$$
\begin{align*}
\begin{bmatrix}
\leftarrow & \ y_{1}^\top & \rightarrow \\
\leftarrow & \ y_{2}^\top & \rightarrow  \\
\leftarrow & \ \vdots & \rightarrow \\
\leftarrow & \ y_{I}^\top & \rightarrow
\end{bmatrix}
& = \begin{bmatrix}
a_{11} & a_{12} & \dots & a_{1R} \\
a_{21} & a_{22} & \dots & a_{2R} \\
 \vdots &  &  & \vdots \\
a_{I1} & a_{I2} & \dots & a_{IR} \\
\end{bmatrix}
\begin{bmatrix}
 \leftarrow &{\ b_{1}}^\top & \rightarrow \\
 \leftarrow &{\ b_2} ^\top & \rightarrow \\
\ & \vdots & \ \\
\leftarrow &{\ b_{R}}^\top& \rightarrow
\end{bmatrix}
\end{align*}
$$

$$
Y = AB,
$$

for matricies $A$ and $B$.

For the factorization to remain interpretable, we need to ensure each row $b_r$ of $B$ is a density. This means we would like to constrain each row $B[r,:]=b_r$ to the simplex^[This assumes the entries $b_r[j]$ in the discretization of the density $b_r$ represent probibilities or areas under some continuous 1D density function, not the sample values of the density $b_r(x_j)$. Assuming $J$ sample points $x_j$ of a grid on the interval $[x_0,x_J]$, the entries of the discretization vector can be defined as $b_r[j]=b_r(x_j) (x_j - x_{j-1})$.]

$$
b_r\in\Delta_J = \left\{ v\in\mathbb{R}^J \middle| \sum_{j=1}^J v[j] =1\quad \text{and}\quad  \forall j\in[J],\ \ v[j] \geq 0 \right\},
$$

which we can write as constraining the matrix $B$ to the simplex

$$
B\in\Delta_{R, J}=\left\{B\in\mathbb{R}^{R\times J} \middle| \forall r\in [R],\, \sum_{j=1}^J B[r,j] = 1  \quad \text{and}\quad \forall (r,j)\in [R]\times [J],\, B[r,j]\geq 0 \right\}.
$$

Given the rows of $B$ represent densities, to ensure the rows of the reconstructed matrix $\hat{Y}=AB$ are still densities, we need the mixing coefficients to be nonnegative $a_{ir}\geq 0$ and rows to sum to one $\sum_{r}a_{ir} = 1$. This constrains the matrix $A$ to the simplex

$$
A\in\Delta_{I, R}=\left\{A\in\mathbb{R}^{I\times R} \middle| \forall i\in [I],\, \sum_{r=1}^R A[i,r] = 1  \quad \text{and}\quad \forall (i,r)\in [I]\times [R],\, A[i,r]\geq 0 \right\}.
$$

<!--
$\newcommand{\blue}[1]{\color{blue}{#1}}\newcommand{\green}[1]{\color{green}{#1}}\newcommand{\purple}[1]{\color{purple}{#1}}\newcommand{\orange}[1]{\color{orange}{#1}}$

$$
\begin{align}
{\mathbf y_1} &= a_{11} \blue{\mathbf b_1} + a_{12} \green{\mathbf b_2} + \dots + a_{1R} \purple{\mathbf b_R} \\
{\mathbf y_2} &= a_{21} \blue{\mathbf b_1} + a_{22} \green{\mathbf b_2} + \dots + a_{2R} \purple{\mathbf b_R} \\
            &\vdots \\
{\mathbf y_I} &= a_{I1} \blue{\mathbf b_1} + a_{I2} \green{\mathbf b_2} + \dots + a_{IR} \purple{\mathbf b_R} \\
\end{align}
$$

$$
\begin{align*}
\begin{bmatrix}
\leftarrow & \mathbf y_{1}^\top & \rightarrow \\
\leftarrow & \mathbf y_{2}^\top & \rightarrow  \\
\leftarrow & \vdots & \rightarrow \\
\leftarrow & \mathbf y_{I}^\top & \rightarrow
\end{bmatrix}
& = \begin{bmatrix}
a_{11} & a_{12} & \dots & a_{1R} \\
a_{21} & a_{22} & \dots & a_{2R} \\
 \vdots &  &  & \vdots \\
a_{I1} & a_{I2} & \dots & a_{IR} \\
\end{bmatrix}
\begin{bmatrix}
 \leftarrow &\blue{\mathbf b_{1}}^\top & \rightarrow \\
 \leftarrow &\green{\mathbf b_2} ^\top & \rightarrow \\
& \vdots & \\
\leftarrow &\purple{\mathbf b_{R}}^\top& \rightarrow
\end{bmatrix}
\end{align*}
$$

-->

## The two approaches for simplex constraints

With constraints being defined in a flexible manner (see @sec-constraints), we decided to test conventional wisdom that Euclidean projections are the right kind of map to use when enforcing a constraint. A constraint that came up in applications was enforcing the $1$st mode slices of a tensor (e.g. rows in a matrix) or $3$rd mode fibres of a third order tensor to lie in their respective simplex.

For example, to constrain a vector $v\in\mathbb{R}^J$ to the simplex

$$
\Delta_J = \left\{ v\in\mathbb{R}^J \middle| \sum_{j=1}^J v[j] =1,\quad \text{and}\quad  \forall j\in[J],\ \ v[j] \geq 0 \right\},
$$

we could apply a Euclidean projection

$$
v \leftarrow \arg\,\min_{u\in\Delta_J} \lVert u - v \rVert _2^2,
$$

or a generalized Kullback-Leibler divergence projection

$$
v \leftarrow \arg\,\min_{u\in\Delta_J} \sum_j u[j] \log{(\frac{u[j]}{v[j]})} - u[j] + v[j]
$$ {#eq-kl-projection}

amoung other reasonable maps onto $\Delta_J$.

The Eucledian simplex projection can be done with the following implimentation of Chen and Ye's algorithm [@chen_projection_2011]. The essence of the algorithm is to efficiently compute the special $t\in \mathbb{R}$ so that

$$
v \leftarrow \max(0, v - t\mathbb{1}) \in \Delta_J.
$$ {#eq-simplex-projection}

The $\max(0, x)$ function should be understood as operating entrywise on $x$. In BlockTensorDecomposition, we use the helper `ReLU(x) = max(0, x)` for this function to assist with broadcasting.

```julia
function projsplx(v)
    J = length(v)

    if J==1 # quick exit for trivial length-1 "vectors" (i.e. scalars)
        return [one(eltype(v))]
    end

    v_sorted = sort(v[:]) # Vectorize/extract input and sort all entries
    j = J - 1
    t = 0 # need to ensure t has scope outside the while loop
    while true
        t = (sum(@view v_sorted[j+1:end]) - 1) / (J-j)
        if t >= v_sorted[j]
            break
        else
            j -= 1
        end

        if j >= 1
            continue
        else # j == 0
            t = (sum(v_sorted) - 1) / J
            break
        end
    end
    return ReLU.(v .- t)
end
```

This is turned into an operation that can mutated `v` with the following definition,

```julia
function projsplx!(y)
    y .= projsplx(y)
end
```

and can be turned into a `ProjectedNormalization` (see @sec-constraints) with the following code.

```julia
simplex! = ProjectedNormalization(isnonnegative_sumtoone, projsplx!)
isnonnegative_sumtoone(x) = all(isnonnegative, x) && sum(x) ≈ 1
```

The generalized Kullback-Leibler divergence projection as stated in @eq-kl-projection is only well-defined when $v[j]>0$ for all $j\in[J]$. In this case the solution is given by

$$
v \leftarrow \frac{v}{\sum_{j} v[j]}
$$
